\chapter*{Acknowledgments}

When I was a kid, I thought every person living on Earth should speak the same language. In my young self's mind, it made perfect sense, because things would be so much easier for everybody. Communication is key after all. As I grew up, however, I realized that language is one of the many cultural aspects that makes us who we are. Humans are unique creatures that are full of dreams, fears, and experiences. I honestly believe that learning from experiences is essential, it can transform the way we live and work. That idea echoed in my mind for years, until the day I was gifted with the opportunity of doing a PhD while being sponsored. I did everything in my reach to make this happen away from the land I call home, Brazil. My homeland has great PhD programs, but I wanted to see things using new cultural lenses. Lenses that are different from the ones I was given at my birthplace. I wanted to work following a mindset that I never experienced before, using a language that is not my own. The PhD portrayed in this book was a life-changing journey for me. I have learned about many technical topics, but I have certainly learned about different cultures and ideas. Before doing my PhD, I had never set foot in Sweden, but now I will never forget the day I landed in Scandinavia. Finally, I must say that my PhD journey was certainly not an isolated event. It was the culmination of several steps that were laid out throughout years and were influenced by many people. I could fill pages of this thesis with names I would like to thank, even knowing that I incur the danger of leaving someone out. Below is my tentative of saying thank you to all those people, in no particular order.

\vspace{7pt}\hrule\vspace{5pt}

A heartfelt, warm and special thank you to \textit{Marilia Landerdahl Abreu}, my wife, who stood with me through good and bad moments in life. Your love, dedication and support are always paramount in my life. I will never forget your courage and sacrifices towards this PhD. Thank you for accompanying and helping me in this great journey, which was unknown, far and so different from what we were used to. I am very thankful for having you and your love by my side. My love for you gives meaning to even the smallest things.

\vspace{7pt}\hrule\vspace{5pt}

I would like to state a huge thank you to two persons that directly guided my work and contributed to its completion: my supervisors \textit{Per Backlund} and \textit{Henrik Engstr{\"o}m}. A significant part of having a great experience during a PhD is all about good supervision, which I had plenty of. My supervisors not only encouraged me, refined my ideas, respected my opinions and allowed me to become academically independent, they have taught me how to do proper research. They worked beyond the duties of supervision to ensure I had a great staying in Sweden, showing me places to visit, things to do and try. I also need to mention how well they helped me deal with many situations outside my PhD that impacted my work somehow, including my doubts and anxieties.

\vspace{7pt}\hrule\vspace{5pt}

I would like to express my appreciation and gratitude towards the \textit{Federal University of Fronteira Sul (UFFS)} and \textit{Conselho Nacional de Desenvolvimento Cient\'{i}fico e Tecnol\'{o}gico (CNPq)}, the two Brazilian government institutions that funded my PhD. UFFS, my employer, allowed me to pursue a PhD while continue paying my regular salary in full, even if I was not there to fulfill my duties. UFFS is a young institution with less than 10 years of history when I left for my PhD, but it never prevented the investment and trust in me. I also want to thank my UFFS colleagues and faculty members that in any form helped me with my PhD. I extend my thank you to CNPq, which also deposited trust in me by approving my project for a four-year scholarship. The funding paid for all my travel expenses to move in and out of Sweden, as well as complemented my financial support there. Finally, I want to thank the \textit{University of Sk\"ovde} and the \textit{EU Interreg \"OKS project Game Hub Scandinavia} for sponsoring my activities as well.

\vspace{7pt}\hrule\vspace{5pt}

Thank you to \textit{Bj{\"o}rn Berg Marklund} for the friendship and for welcoming me into his family. Your efforts helped me find the way in the Swedish system in several matters. I have learned so much about the Swedish culture while having an awesome time barbecuing, game jamming, playing \textit{Contra III}, petting Digby, or having a nice day by the lake (among many others). Thank you to \textit{Adriano Costa} and his family for the friendship and great moments together. I appreciate you filming my spectacular falls while (trying) to ice skate, having dinners or helping me when heavy clouds were upon. As I once said, it is amazing to find the best Brazil has to offer outside Brazil itself. Finally, thank you to \textit{Denio Duarte} and his family for great moments together as well. I bet we will never forget the $-35^{\circ}$C experienced in north Sweden, or the walks in the woods of Varola.

\vspace{7pt}\hrule\vspace{5pt}

I had the pleasure of working and sharing moments with great people from my department and from the university. Thank you all! Your help towards me, my studies and my experiments was invaluable. A special thank you to \textit{Torbj{\"o}rn Svensson} for helping with cameras and lights, as well as for welcoming me in such a nice way right after my arrival in Sweden. I will never forget the tour around Varnhem and my very first Swedish fika shared with your lovely family. Thank you also to \textit{Marcus Toftedahl} for all the help, including finding subjects for my experiment, gearing up my data processing, or teaching me how to cook mushrooms. I also want to thank \textit{Katrin Dannberg} for all the time you put in helping me with my experiment, in particular spreading the word about my study. I could never have found so many subjects without your help. I enjoyed all the Swedish you have taught me, even if I managed to learn just a few words. Thank you to \textit{Mikael Lebram} for helping me gear up my experiments and for showing me (and let me try) so many cool projects, like the car simulator. Finally, thank you to \textit{Andrea Di{\~a}o \-Jonsson} for helping me with the paperwork I had to go through to make this PhD happen, for helping me land in Sweden with a place to stay, and for all the help regarding my experiments.

\vspace{7pt}\hrule\vspace{5pt}

I want to acknowledge and thank the members of my defense committee and opponents who invested their time and expertise to ensure the quality of this work. I also want to note the contributions from those who evaluated my work since its proposal until its final form: a big thank you to \textit{David Vernon}, \textit{Mikael Johannesson}, \textit{Stefan Seipel}, and \textit{Veronica Sundstedt} for the invaluable feedback regarding my research. Your comments helped me steer my research towards fruitful venues, which led me to produce a better piece of work. I want to thank as well all those teaching in the courses I attended at the University of Sk{\"o}vde. I have learned a great deal in those classes regarding a variety of topics and methods. Particularly, I want to thank \textit{Jeremy Rose} for the essential help when I was selecting theories to support my thesis, and also for lowering my own positivist and quantitative walls, allowing me to see the bright light of the philosophy of science. Thank you to \textit{Cesar Tadeu Pozzer}, who made me aware of the University of Sk{\"o}vde and its PhD program years after being my supervisor during my masters. You provided me with essential information to enroll in the program and to move to Sweden, igniting this whole journey. Finally, I want to thank my professors at UFSM, particularly \textit{Andrea Schwertner Char{\~a}o} and \textit{Marcos Cordeiro d'Ornellas}. You all prepared me for the great challenges ahead of any undergraduate program, which helped me conduct my PhD work several years after we have crossed paths.

\vspace{7pt}\hrule\vspace{5pt}

Thank you to \textit{Antonieta}, \textit{Delcio}, and \textit{Gabriele Bevilacqua}, my mother, father, and sister. You always surrounded me with love and were there for me in regard to all sorts of matters, including education and personal development. I would never be here without your solid foundation. Life has winds of its own and we sometimes face difficult and heavy decisions, but I am sure we all have learned from those. I appreciate your love and thoughts towards me.

\vspace{7pt}\hrule\vspace{5pt}

Doing a PhD can be a lonely experience, but thankfully I had an awesome group of fellow PhD students around me. You all helped me advance my own studies in a way or another, be it with a quick talk, a word of advice, time to participate in my experiments, discussing ideas, or by simply sharing common struggles so I could put my own problems into perspective. A huge thank you to all PhD students at Portalen from the second to the fifth floor. A particular thank you to the guys I regularly met (at $3\pm5$) during the fika time on the fifth floor: \textit{Elio Ventocilla}, \textit{Navoda Senavirathne}, \textit{Niclas St{\aa}hl}, \textit{Andr{\'a}s M{\'a}rki} and \textit{Nikolas Huhnstock}. Sharing joy, our daily fights, or comparing the size of a sequoia tree to a building were important moments for me to keep things grounded and sane.

\vspace{7pt}\hrule\vspace{5pt}

Finally, I want to make a special note of appreciation for those who direct or indirectly contributed to my work: all participants of my experiments, \textit{Vera Lindroos} for proofreading this thesis, \textit{Thomas Fischer} for creating and helping me with the {\LaTeX} template that typesets this thesis, \textit{Radu Dinu}, \textit{Elin Tomasdottir}, \textit{Anette Andersson}, \textit{Melissa Mozifian}, and my SFI teachers: \textit{Marianne Persson}, \textit{Patrik Fred{\'e}n de los Rios} and \textit{Ulf \-Nilsson}. Last, but in absolutely no circumstance least, thank you to \textit{Espresso House} and its great coffee and friendly staff. Your coffee fueled my work and this thesis to the point I didn't even have to specify my order to your baristas; they already knew what I wanted. If this thesis smells anything, it smells \textit{``a salted caramel latte with cream to take away, please"}.

\vspace*{\fill}

\textit{This work has been performed with support from: CNPq, Conselho Nacional de Desenvolvimento Cient\'{i}fico e Tecnol\'{o}gico - Brasil; University of Sk\"ovde; EU Interreg \"OKS project Game Hub Scandinavia; UFFS, Federal University of Fronteira Sul.}
